A introduction to lattice trapdoor

### 2023-11-26 co687/project/main.tex
Use NearestPlane to implement SizeReduce

\subsection{LLL lattice basis reduction algorithm}
At the time of this survey, the best algorithm for solving the (approximate) shortest vector problem is the LLL lattice basis reduction algorithm, attributed to Arjen Lenstra, Hendrik Lenstra, and László Lovász. The reduction algorithm transforms an input basis into an LLL-reduced form, where the first base vector of the reduced basis is an approximation of the shortest vector.

If time permits, we will discuss the algorithm itself. However, I found it okay to first treat the algorithm as a black box, and instead focus on the properties of the reduced basis.

@@

- For all $j > i$, $\vert\mu_{j, i}\vert \leq \frac{1}{2}$
- For all $1 \leq i \leq n-1$, $\delta\Vert \pi_i(\mathbf{b}_i)\Vert^2 \leq \Vert\pi_i(\mathbf{b}_{i+1})\Vert^2$

@@

Also recall the Gram-Schmid orthogonalization process:

@@

@@

One of the earliest instances of such class of lattice trapdoor is proposed by Goldreich, GoldWasser, and Halevi in their 1997 paper \textit{"Public-key cryptosystem from lattice reduction problem"}. At a high level, the trapdoor is parameterized by three items: a "bad" basis $B$, a "good" basis $R$, and an error bound $\sigma$. In the forward direction, the function maps a pair of lattice coordinate $\mathbf{v} \in \mathbb{Z}^n$ and a small error vector $\mathbf{e} \leftarrow \{ -\sigma, \sigma \}^n$ to $\mathbf{x} = B\mathbf{v} + \mathbf{e} \in \mathbb{R}^n$. If the parameters are generated correctly, then the closest lattice point in $\mathcal{L}(B)$ is exactly $B\mathbf{v}$.

>>>>

\subsection{Reduced lattice basis}
At the time of this survey, the best algorithm for solving the (approximate) shortest vector problem is the LLL lattice basis reduction algorithm, attributed to Arjen Lenstra, Hendrik Lenstra, and László Lovász. The reduction algorithm transforms an input basis into an LLL-reduced form, where the first base vector of the reduced basis is an approximation of the shortest vector. Details of the actual algorithm for obtaining a reduced basis will be discussed in section 4. Meanwhile, it is helpful to first discuss the properties of the reduced form itself.

@@

\begin{enumerate}
    \item For all $j > i$, $\vert\mu_{j, i}\vert \leq \frac{1}{2}$
    \item For all $1 \leq i \leq n-1$, $\delta\Vert \pi_i(\mathbf{b}_i)\Vert^2 \leq \Vert\pi_i(\mathbf{b}_{i+1})\Vert^2$
\end{enumerate}

@@

Also recall the Gram-Schmidt orthogonalization process:

@@

\subsection{Basis size reduction}
In addition, to being the best known algorithm for solving the CVP problem, Babai's nearest plane algorithm is also an essential component to the LLL lattice basis reduction algorithm. Specifically, the nearest plane algorithm is used to reduce the size of the basis vector so the first condition of the reduced basis (shown below) can be satisfied:

$$
\forall j<i, \vert\mu_{i, j}\vert \leq \frac{1}{2}
$$

The size reduction algorithm, which we will denote by $\operatorname{SizeReduce}$, takes a basis $B$ and returns a size-reduced basis $B^\prime$ such that the condition above is true. The procedure for $\operatorname{SizeReduce}$ is as follows:

For each of $i \in \{2, 3, \ldots, n\}$
\begin{enumerate}
    \item $\mathbf{v} \in \mathcal{L}([\mathbf{b}_1, \ldots, \mathbf{b}_{i-1}]) \leftarrow \operatorname{NearestPlane}([\mathbf{b}_1, \ldots, \mathbf{b}_{i-1}], \mathbf{b}_i - \mathbf{b}_i^\ast)$
    \item substitute $\mathbf{b}_i$ with $\mathbf{b}_i - \mathbf{v}$, then update $B$ and $B^\ast$
\end{enumerate}

Recall the result of the nearest plane algorithm we know that:

$$
\forall j < i,
\frac{
    \langle
        \mathbf{b}_i - \mathbf{b}_i^\ast - \mathbf{v}, \mathbf{b}_j^\ast
    \rangle
}{
    \langle 
        \mathbf{b}_j^\ast, \mathbf{b}_j^\ast
    \rangle
} \in [-\frac{1}{2}, \frac{1}{2})
$$

Notice in the equation above, we have $\mathbf{b}_i^\ast \perp \mathbf{b}_j^\ast$ because $j < i$. We also have $\mathbf{b}_i - \mathbf{v}$ being the new value for $\mathbf{v}_i$ after the substitution. This means that after the substitution:

$$
\forall j < i,
\frac{
    \langle
        \mathbf{b}_i, \mathbf{b}_j^\ast
    \rangle
}{
    \langle
        \mathbf{b}_j^\ast, \mathbf{b}_j^\ast
    \rangle
} \in [-\frac{1}{2}, \frac{1}{2})
$$

The LHS of the equation above is exactly $\mu_{i,j}$. Thus we have satisfied the first condition of $\delta$-LLL reduced basis. In addition, because $\mathbf{v} \in \mathcal{L}([\mathbf{b}_1, \ldots, \mathbf{b}_{i-1}])$, the substitution $\mathbf{b}_i \leftarrow \mathbf{b}_i - \mathbf{v}$ is equivalent to adding onto the substituted basis a linear combination of other basis vectors, which corresponds to a unimodular matrix multiplication and thus does not change the lattice itself before or after the size reduction.

\subsection{The LLL basis reduction algorithm}
By applying $\operatorname{SizeReduce}$ to the basis, we can obtain a size-reduced basis that satisfy the first condition of $\delta$-LLL reduced form, but there is no guarantee that the second condition will hold. Pairs of adjacent basis that violate the second condition can be swapped so that the swapped pair does not violate the second condition. Because swapping columns is equivalent to right multiply by unimodular matrix, the lattice it preserved, but the first condition will not hold, so we will need to apply $\operatorname{SizeReduce}$ again.

The LLL basis reduction algorithm is exactly a repetition of alternating between size reduction and column swapping. For $\frac{1}{4} < \delta < 1$, it can be proved that the number of iteration is bounded by some polynomial factors of the dimension $n$ of the lattice. For $\delta = 1$ we will obtain the best possible reduction, but there is no guarantee that the basis reduction will run in polynomial time.

@@

One of the earliest instances of such class of lattice trapdoor is proposed by Goldreich, Goldwasser, and Halevi in their 1997 paper \textit{"Public-key cryptosystem from lattice reduction problem"}. At a high level, the trapdoor is parameterized by three items: a "bad" basis $B$, a "good" basis $R$, and an error bound $\sigma$. In the forward direction, the function maps a pair of lattice coordinate $\mathbf{v} \in \mathbb{Z}^n$ and a small error vector $\mathbf{e} \leftarrow \{ -\sigma, \sigma \}^n$ to $\mathbf{x} = B\mathbf{v} + \mathbf{e} \in \mathbb{R}^n$. If the parameters are generated correctly, then the closest lattice point in $\mathcal{L}(B)$ is exactly $B\mathbf{v}$.
